3DPAS: Distributed Dynamic Data-intensive Programming Abstractions and Systems

### 2015-05-29 old-sections/3dpas_public_paper.tex
adding other, mostly old material

\author{ Shantenu Jha, J.D. Blower, Neil Chue-Hong, \\[-0.1em]
Simon Dobson, Daniel S. Katz, Omer Rana}

@@

\begin{abstract}
  Many problems at the forefront of science, engineering, medicine,
  and the social sciences, are increasingly complex and
  interdisciplinary due to the plethora of data sources and
  computational methods available today.  A common feature across many
  of these problem domains is the amount and diversity of data and
  computation that must be integrated to yield insights. Such data is
  increasingly large-scale and distributed;
  the data-sets, their location and scheduling decision
  are time-dependent. % Applications need access
  % to these disparate data-sets and to the methods that operate on
  % them.  Data-intensive applications need first-class support for
  % distributed heterogeneous platform execution.
  The 3DPAS Research theme (funded by the eSI, Edinburgh and US NSF),
  aims to investigate applications that have these characteristics, in
  other words it operates at the triple point of {\it dynamic} and
  {\it distributed} and {\it data-intensive}, (3D) attributes.  This
  evolving document represents work in progress of the 3DPAS Research
  Theme.
\end{abstract}

@@

@@

\jhanote{Overall introduction, leading into some specific examples.}

@@

Why bother about distributed, dynamic and data-intensive applications?
\jhanote{(Shantenu, Omer, Neil, Simon, and Dan)}

@@

There is a general appreciation that the challenges of data-intensive
applications span beyond large-volumes and management/curation.  One
specific attribute that we believe is pervasive but not necessarily
obvious is “dynamic data”. The belief is based upon the fact that data
is often distributed, and as a consequence many attributes of
distributed systems permeate to data. Additionally the requirement of
many applications impose a requirement that data is dynamic.

\subsection{Types of Dynamic Data}
Some example of such dynamism:

\begin{itemize}
\item Change of data operated upon.
\item Variability of data rate -- arrival, consumption or ability to generate (due to dependencies)
\item Data placement and scheduling issues are time dependent
\item The nature of coordination between the different execution units
changes
\end{itemize}

\jhanote{Possibly merge?}

\begin{enumerate}

\item Streaming data -- data comes and goes
  \url{http://traffic.berkeley.edu/} and
  \url{http://lagrange.ce.berkeley.edu/fsn/}
\item Data set that is being operating upon is changing
\item Data has to be transformed
\item Placement decisions time-dependent
\end{enumerate}

@@

\jhanote{Good Q: I think other than scale -- of data volume, and
  degree of variability they are not too dissimilar}

\subsection{Characterising Dynamic Applications}

Amdahl’s number has been useful in characterising “static”
data-intensive applications and systems. We believe there is scope to
generalize Amdahl’s number to capture attributes of dynamic data
applications.

\subsubsection{Dynamic Data in a Web Environment}

\emph{Just some placeholder notes from Jon.  Could belong elsewhere.}

@@

\katznote{new suggested questions from Don Middleton: How much diverse data integration is involved?  how diverse is the data?}

@@

\subsubsection{New answers from Steve}

@@

%\subsubsection{Old answers from Steve}
%
%   1.       What is the purpose of the application?
%
%A physics group is interested in a specific physics channel (underlying physical process) as recorded by the Atlas experiment at the LHC (Large Hadron Collider) at CERN in Geneva. The set of available data grows steadily over time as more data are collected and reconstructed. It would not be efficient to allow individual physics groups to process this large volume of data in an uncoordinated manner as it requires reading all available data.
%
%   2.       How is the application used to do this?
%
%Physics groups register their analysis code to be run on the next scheduled analysis ``train''. The train leaves every two weeks and reads through all the data, runs the physics selection code of each group and writes out the selected events and output from the computation to produce relatively small data sets for each physics group. The groups then use these selections to perform more ad-hoc analysis.
%
%   3.      What infrastructure is used? (including compute, data, network, instruments, etc.)
%
%The WLCG (Worldwide LHC (Large Hadron Collider) Computing Grid) is used with a mixture of gLite software, experiment specific software and physics group specific software. WLCG has 250,000 cores distributed over 140 sites and with 100~PB of disk. There is no wide area dedicated network infrastructure.
%
%   4.      What dynamic data is used in the application?
%
%The data being processed can be considered as dynamic as it grows steadily as new raw data are collected and the basic reconstruction program is run to produce information about specific particles involved in a collision. This reconstruction is rerun periodically (two or three times per year is expected) as calibrations of the detector are improved. Reconstruction algorithms are also improved as are the physics groups' selection code. Calibration data are held on an RDBMS with Frontier/Squid front ends.
%
%         a           What are the types of data,
%
%          The bulk of the data is individual event data at different stages of refinement. The is held in a non-standard representation of C++ serialized objects.
%
%         b.             What is the size of the data set(s)?
%
%          Currently the reconstructed data (excluding copies) is about m TB and is growing at about  n TB per month. Many data files are replicated on multiple sites. This is done both by copying it to where it is likely to be useful and partly dynamically as the need arises.
%
%   5.       How does the application get the data?
%
%This is mostly by copying the file to where the job will run - though some remote file access is also used.
%
%   6.  What are the time (or quality) constraints on the application?
%
%If some data are missed during processing the only effect is a reduction of statistical precision in the final analysis. If the selection code has been improved then naturally the physics groups would like to see the results as soon as possible.
%
%
\subsection{Astrophysics -- Virtual Observatory (Bob) \label{astroBob}}
\begin{itemize}
\item Analysis of data to generate event streams $\rightarrow$ lead to the
  positioning of telescopes
\end{itemize}

@@

I'll assume that the ``application'' in question is the sum of the two
activities, despite the fact that they are different in many respects,
and I'll try and distinguish between them where appropriate.

Let me know if you want clarification or further information - in some cases I wasn't quite sure what you were after.

@@

For the sea mammals the animal crawls up a beach, the sensor packages
sees the cellular network and transmits. For environmental missions
we're still wondering about that ourselves :-)

@@

Not sure I see what you mean here. Generally it's just a matter of
capturing at whatever resolution device the sensor pack can support.

I could probably get more complete -- perhaps *way* too complete :-) --
answers from Mike Fedak in Scottish Oceans if that'd help.

%\subsection{Bioinformatics Applications (Tom Slezak) - do not publish \label{bioTom}}
%
%1. What is the science/engineering problem? What is the purpose of the application?
%
%Comparing metagenomic (DNA sequence) data sets against each other. This could be to distinguish human disease biomarkers among different individuals, determine different soil (or water or air or ...) microbial composition from different samples, etc.
%
%2. How will the application solve the problem?
%
%In general this is a large (for biology -- don't think in terms of exascale!) data problem with large (all-vs-all comparison) computing. There are strategies to make this more scalable that will reduce the computing volume.
%
%(a) Which of the following describes the application
%
%         i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)
%
%         ii) large (exascale) data and modest computing (e.g., search)
%
%         iii) modest data and large (exascale) computing (e.g., simulation/modeling)
%
%(b) Can you quantify the amount of computing vs the amount of data?
%
%Potentially peta-bytes or exa-bytes of data. Not sure how to quantify the computing. (Huge memories will be potentially more valuable than huge numbers of cores -- although this is heresy to the physics crowd.)
%
%3. What infrastructure will be used? (including compute, data, network, instruments, etc.)
%
%(a) What will be distributed?  data, compute, both?
%
%Individual data sets will be terabyte-scale. There could be many thousands of such sets to compare. Having them all local may not be feasible (the transfer time from where they are generated by others may be prohibitive. Computing might then have to be distributed. Biology is wrestling with this now.
%
%4. What data be used in the application?
%
%(a) What are the types of data
%
%Genomic sequence, billions of short (35-100?) bases of DNA; unassembled
%
%(b) What is the size of the data set(s)?
%
%Terabyte scale
%
%(c) Will any data be dynamic?
%
%Not for this problem
%
%5. How will the application get the data?
%
%That is a major part of the problem! Data will (in general) be generated from other places around the world. Getting to the data will be an issue.
%
%6. Are there time (or quality) constraints on the application?  (If so, what are they?)
%
%You'd like answers in under a day, generally.
%
%7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.
%
%Biology is wrestling now with  about 3 orders of magnitude increase in DNA sequence data generation capability in the past 2 years. All algorithms, architectures, storage, and networks have broken or will break soon. Some biology problems (not necessarily the one I describe here) are not good fits for core-rich and memory-poor modern architectures. For some problems (comparing k-mers across metagenomic data sets, for instance) you need terabyte-scale memory, not thousands of cores.
%

@@

\subsubsection{CDAT} (from Dan)

The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a scriptable
(Python) set of packages and tools that are currently heavily used for
analysis of climate data.  CDAT has been developed by LLNL's Program
for Climate Model Diagnosis and Intercomparison (PCMDI), and is often
used with data from the Earth System Grid (ESG)~\cite{ESG}.  The most
common current usage of these tools with the data is that a user
downloads a set of data from ESG to a local resource, then uses the
CDAT to analyze the data.  However, the PCMDI group is now planning to
support a new usage mode, where data transfer can be scripted.
Initial development of this capability is planned for February 2011,
using the Globus online data transfer service~\cite{globusOnline} as a
scriptable component.

\subsubsection{ESGF} (from Don and Dan)

Aimed at supporting international CMIP/IPCC intercomparison activity.
Three stages:

{\bf Data Acquisition (Modeling)}

\begin{itemize}
\item Climate centers run prescribed set of common experiments; produce 2-10 PB of data
\begin{itemize}
\item data can also come from sensors - here, the data would still be post-processed before it would be published.
\end{itemize}
\item Centers can publish their own output data or send it to another center to publish data generated over \~{}2 year period (then post-processed and published over a few more months)
\item Properties: data-intensive, distributed centers, lots of computing
\end{itemize}

{\bf Data storage (Infrastructure/system)}

\begin{itemize}
\item ESGF develops and deploys a federated network of gateways and associated data nodes
\item As models run at each center, output data is post-processed into packages with common formats and prescribed metadata conventions
\item Most centers will deploy data node software stack, and using this, they manage the data from their experiments
  \begin{itemize}
  \item the data node software stack scans the data, makes sure it has the right metadata fields, does QA/QC, build a set of catalogs of the prepared data
  \begin{itemize}
                \item minimally, catalog provides HTTP links to data elements
                \item also can provide GridFTP endpoints
                \item also can provide product services that abstract the dataset in other ways - get whole file, subset, browse, etc.
  \end{itemize}
  \end{itemize}
    \item When the center is happy with the data/catalog in the data node, they publish it to a host/affiliated gateway
\begin{itemize}
         \item  this submits the catalog to the gateway
          \item  them the gateway shares this catalog with other gateways so that all gateways have a consistent view of all the published data
\end{itemize}
    \item  Core archive (1-2 PB, most popular data)
\begin{itemize}
          \item  will be replicated at several sites
         \item  replication activity manually requested/initiated by a gateway owner
          \item  replication: copies catalog to replica site, do data transfer to new data node, then publish  catalog, push catalog to (this) gateway associated with new data node
\end{itemize}
    \item  Currently building notification service
\begin{itemize}
          \item  track which users have downloaded data - if a problem is found in a data set or if new data is added to a data set, notify users who have downloaded/replicated it
\end{itemize}
    \item  Properties: data-intensive, distributed gateways and data nodes, data appears in the system over time (dynamic)
\end{itemize}

{\bf Data processing (Application)}

\begin{itemize}
    \item  Users (\~{}20000 as of Dec. 2010) can:
\begin{itemize}
         \item  browse/search a catalog at any gateway and locate data, which might be hosted by a data node affiliated with another gateway. (can output a wget script that can later fetch the data)
         \item  authenticate, gain access to a group
        \item  download data via http or GridFTP or access product services (uses data retrieval syntax (DRS) - can be scripted)
  \end{itemize}
\end{itemize}

\begin{itemize}
    \item  abstractions that can be used by a programmer: data is stored in NetCDF format, uses CF metadata conventions, can be accessed using DRS
    \item  Some users will analyze data from a single model
    \item  Many applications are multi-model analyses - many users want to look at the same parts of the output of some/all of the models, to understand if and how the models differ
    \item  Some centers will gather some/all of the core archive (plus more, perhaps) on local systems for local users to perform ``power analysis''
\end{itemize}

\subsection{Astro (LSST) (Adam Barker) \label{astroAdam}}

\begin{enumerate}
\item What is the science/engineering problem? What is the purpose of the application?

Find and study variable objects and moving objects.

\item How will the application solve the problem?

Processing of data (data analysis and detection of new sources) from each image should be done while the next image is being taken by the telescope.  This means that if anything unusual is detected, normal observation can be interrupted. Other observing resources can also be notified instantly, so they can observe the same event. As data are collected, they will be added to all the data previously detected from the same location of sky to create a very deep master image. LSST will also build up a database of all known moving objects.

Every time a new image of the sky is obtained, the master image will be subtracted from it. The result is a subtracted image that contains the difference between the current sky image and its average state.

This subtracted image is then processed by a cluster of computers with the following steps:
\begin{enumerate}

\item Query object catalog (knows orbits of all known objects) to find objects which are expected to appear in subtracted image, given the area of sky, time of day.

\item Cross-match expected objects with sources in subtracted image. The result is the unmatched-source catalog, which contains sources that canÕt be matched with a previously known object

\item Use unmatched-source catalogue to compute pairs of detections separated by short time intervals, called tracklets. A tracklet is an observable, short section of orbit. In order to create pairs of detections, all objects in the unmatched-source catalog are queried against the orbit catalog, in an attempt to obtain data about these objects from earlier observations. If earlier observations and current observations can be linked to form tracklets, they are stored in the tracklet catalog.

\item Use tracklet catalog to attempt to link these tracklets over a larger time window. This is achieved by again querying the orbit catalog, this time looking for observations of the same objects even further back in time. If matches can be found and the tracklets can be extrapolated out into longer sections of orbit, then they are added as new orbits to the orbit catalog.

\item Update orbit catalog with redetections of known objects. Each rediscovery provides information that improves the known orbits.

\item Attempt to classify all entries in the orbit catalog (i.e., all the objects which now have known orbits). If any Near Earth Objects are detected to be passing close to the earth, alerts are generated to astronomers so that follow up observations can be scheduled.

\item Compare with other observatories: If an unknown object is found that cannot be classified locally
\begin{enumerate}
\item issue call for participation to set of possible observatories
\item each observatory accepts or rejects the call
\item local observatory chooses from accepting observatories one to obtain data from
\item that observatory takes data and sends it to local observatory
\item local observatory decides what to do - classify the object, contact a human, call for more observations (automatically to other observatories, that decide whether or not to accept, and when)
\end{enumerate}

\end{enumerate}

\begin{enumerate}
\item Which of the following describes the application

large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)

\item Can you quantify the amount of computing vs the amount of data?

LSST will generate 36~GB of data every 30 seconds, and over a 10-hour winter night, will collect up to 30~TB.

\end{enumerate}

\item What infrastructure will be used? (including compute, data, network, instruments, etc.)

A linked network of observatories and data and computing resources.

(a) What will be distributed?  data, compute, both?

both

\item What data be used in the application?

Fits images, and large databases for analysis of the discoveries.

 (a) What are the types of data

 (c) Will any data be dynamic?

\item How will the application get the data?

Some will be analyzed at the mountain top, discoveries will be sent to other centers.

\item Are there time (or quality) constraints on the application?  (If so, what are they?)

Time component is critical. As real-time as possible.

\item Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.

\end{enumerate}

also, the Palomar Transient Factory is the best example to date:

See:

\url{http://www.astro.caltech.edu/ptf/}

\url{http://xxx.lanl.gov/abs/0906.5350}

\url{http://xxx.lanl.gov/abs/0906.5355}

\url{http://arxiv.org/abs/1011.2199}

\subsection{Interactive exploration of environmental data (Jon) \label{envJon}}
Synopsis of chapter we're writing for Malcolm's book.  Looking at needs of interactive applications as opposed to batch ones.

@@

The applications often take the form of graphical tools that share a lot in common with Geographic Information Systems (GIS). However these applications are much better suited to scientific data, which are large and multidimensional.  The general ÒvisionÓ is of a map-based interface, onto which different datasets can be overlain.  A processing step may be initiated by rubber-banding an area of the map and selecting from a list of algorithms that process data within the selected area, perhaps calculating statistics of the data.

@@

Usually the applications are driven by data served through web services of various kinds.  These web services may be ÒfedÓ by data coming from instruments or from computer simulations.  A key challenge is how to integrate compute services with these data services in an efficient, easy-to-use and transparent way, allowing fast data processing.  The processing algorithms are generally quite simple and so it is frustrating to succumb to the high overhead of some scheduling systems.  The infrastructure probably has more in common with the Web (or perhaps cloud) than the Grid.

@@

Dynamic data are only sometimes used, but may arise from Òreal-timeÓ feeds from instruments or from looking at ÒliveÓ results from a model.  Data from instruments are ÒsmallÓ 1MB-100MB but model output may be very big (GB-TB).

@@

% \subsection{Astro (CMB) (Julian Borrill) - do not publish \label{astroJulian}}

% 1. What is the science/engineering problem? What is the purpose of the application?

% Cosmic Microwave Background (CMB) data simulation and analysis. The CMB is an image of the Universe as it was only 400,000 years after the Big Bang. Tiny fluctuations in the CMB temperature and polarization encode the fundamental parameters of cosmology and, using the Big Bang as the ultimate particle accelerator, ultra-high energy physics. Extracting this information from the data gathered by current and anticipated CMB observations is an extremely computationally intensive endeavour for which we have developed our massively parallel tools.

% 2. How will the application solve the problem?

% The central computing challenge for any CMB dataset is the simulation and analysis of O($10^4$) synthetic observations, used to correct for biases and quantify uncertainties in the analysis of the real data. We use preconditioned conjugate gradient techniques to solve for the maximum likelihood sky map given the input data (obtained by scanning the sky with hundreds to thousands of detectors for weeks to years) and its piecewise stationary noise statistics. To avoid the IO bottleneck inherent in the traditional simulate/write/read/map paradigm, all simulations are performed on the fly only when requested by the map-making code.

% (a) Which of the following describes the application

%      i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)

%      ii) large (exascale) data and modest computing (e.g., search)

%      iii) modest data and large (exascale) computing (e.g., simulation/modeling)

% (iii) O(1 - 10) TB input data and a similar volume output.

% (b) Can you quantify the amount of computing vs the amount of data?

% The computing scales with the number of observation samples, while the communication and IO scale with the number of pixels in the map. For the current Planck satellite experiment these are O($10^12$) and O($10^8$) respectively; for the proposed CMBpol satellite they will be O($10^15$) and O($10^8$) - that is, we expect a 1000x increase in calculation cost over the next 15 years (including intermediate suborbital pathfinder experiments) but fixed communication/IO costs.

% 3. What infrastructure will be used? (including compute, data, network, instruments, etc.)

% We are targeting the largest supercomputers available - Hopper, Tianhe, Blue Waters, etc. It's all about the cycles, although as the concurrency we work at increases previously solved IO and communication scaling bottlenecks typically re-emerge.

% (a) What will be distributed?  data, compute, both?

% Both.

% 4. What data will be used in the application?

% Real and simulated CMB data.

% (a) What are the types of data

% (b) What is the size of the data set(s)?

% (c) Will any data be dynamic?

% See above.

% 5. How will the application get the data?

% Real data are read from disk; simulated data are generated on the fly.

% 6. Are there time (or quality) constraints on the application?  (If so, what are they?)

% Typically we want to be able to simulate and map O($10^2$) realizations of an experiment's data in O(1) wallclock hour to provide 10\% errors during the early analysis stages, and O($10^4$) realizations in O(100) wallclock hours for the final definitive analysis.

% 7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.

% Moving to the current and future generations of heterogeneous many-core systems will require a significant re-working of our analysis codes, both to deal with communication over hundreds of thousands of cores and to take advantage of accelerator opportunities such as GPUs.

% \subsection{Fusion (Scott Klasky) -- do not publish \label{fusionScott}}

% 1. What is the science/engineering problem? What is the purpose of the
% application?

% With ITER scheduled to operate before 2020, plasma fusion has become
% one of the highest priority research items for DOE. Scalability of
% first-principles fusion codes is an important part of fusion research
% needed to obtain predictive capability of plasma performance with
% first-principles physics fidelity. We want to build a realistic
% full-distribution kinetic code, in realistic ITER geometry, combining
% the strengths of each of the existing gyrokinetic codes (GEM, GTS,
% XGC, and GTC), and utilize the most up-to-date computer science,
% hardware design and applied mathematics technologies developed. The
% application that we will develop will be capable of 1) discovering
% Macroscopic Magneto Hydro Dynamics activity in burning plasmas, 2)
% Understand Energetic particle effects in fusion reactors, 3)
% Understand the heating in a fusion reactor, and 4) Understand the
% interaction of the plasma at the material boundary.

% 2. How will the application solve the problem?

% (a) Which of the following describes the application

%         i) large (exascale) data and large (exascale) computing (e.g., data mining, analytics, bioinformatics)

%         ii) large (exascale) data and modest computing (e.g., search)

%         iii) modest data and large (exascale) computing (e.g., simulation/modeling)

% The applications we currently run (XGC1, XGCP, GTC, GTS) run on all of the leadership class facilities, and require very large resources for the modest problems which we are currently solving. In the next generation (exascale) we will not only need all of the computing power that will be available at the time to understand ITER, we will also need to understand all of the data that will be generated during the simulation. Our techniques have been to examine and understand the data in situ, to reduce the total amount of data produced in the simulation. Today we have produced O(100 TB/week) for a simulation using our ADIOS middleware incorporated in our simulations. Using our staging techniques, we believe that we will only produce O(10 PB) on an exascale machine. (I don't know if this is considered modest dat, or extreme size data)?

% (b) Can you quantify the amount of computing vs the amount of data?

% 1 PB of data -- 10 PB of data, for an exascale size computation run in O(100 hours).
% Today we achieve about 6\% of peak with excellent scalability to 220K
% cores on the Cray XT systems. It is
% difficult to predict how we will run on
% accelerators, and next generation machines.

% 3. What infrastructure will be used? (including compute, data,
% network, instruments, etc.)

% For Validation, we will need to compare our data from the computer to
% ITER (or current fusion reactors). Typically the amount of data used
% in comparison is tiny by comparison (people only look at MB's form
% experiments).

% (a) What will be distributed?  data, compute, both?

% We always run on all available resources, so if we have computer time
% across the country, then our data will be distributed, and so will our
% computation. ITER will have data that will be distributed, but it
% still unknown.

% 4. What data be used in the application?

% I don't understand this question? We will input data (from perhaps a restart), and this data is typically in the ADIOS-BP file format for large data. Smaller data sets are typically stored in HDF5, and the smallest are stored in MDS+.

% (a) What are the types of data

% (b) What is the size of the data set(s)?

% Depends on what we want to understand --- For GTS/GTC we have typically
% looked at the particle data to compute fluxes, etc. and this data can
% be large, but due to the Monte Carlo error, and ways to generated 5D
% distribution functions from the particles, we reduce the particles to
% a manageable amount (300 GB per time step x 1 K time steps). Our
% current I/O is capable of reading this in $<$ 100 s per time step, which
% allows us to analyze the data in about 1-2 days.

% (c) Will any data be dynamic?

% Don't understand the question. Features change and are dynamic. The physics will get more complex as we continue to evolve our codes, and will be more dynamic in nature.

% 5. How will the application get the data?

% Reading the data from disk --- (Perhaps I don't understand the question).  It will also couple to other codes, and the coupling will be in-memory data exchange.

% 6. Are there time (or quality) constraints on the application?  (If so, what are they?)

% We always need to run as fast as possible since we have limited computing time (our team does have about 100M hours FY 11, but this still puts severe restrictions on what we run, so the faster we can run, the more science we can get done.

% 7. Please feel free to also talk about the current state of the application, if it exists today, and any specific gaps that you know will need to be overcome.

% Our biggest gaps will be in accelerating the code on next generation exascale platforms, and in processing as much data as possible in situ.

\subsection{SOA Astronomy applications from Adam\label{astroSOAAdam}}

Calculates the photometric redshift for a given area of sky [RA, DEC] using two tools: HyperZ and ANNz, both accessible via WS interfaces.

The application is a service-oriented workflow. Services are orchestrated through a pipeline in order to compute the redshift of a given area of sky.

The Wide Field Survey Archive (WFS):  image catalogue
Spectroscopic database: used to provide spectroscopic data for an area of sky

SExtractor tool: image processing, extracts all objects of interest (stars, galaxies etc.)
Cross matching tool: compiles one table of all objects of interest over the 5 wavebands

Hyperz: photometric redshift estimation algorithm 1
ANNz: photometric redshift estimation algorithm 2

mySpace: AstroGrid storage service

\katznote{checking more on this point}

The WFS catalogue will be updated regularly, therefore a query at time x will not provide the same data as a query at time y.

Data sets (images) will be in the GB range, I can attempt to clarify these details.

WS interfaces.

Enough data needs to be available for a given area of sky from both the WFS archive and the spectroscopic archive.

\subsection{TNO + Thales (Netherlands)  -- Omer (Kees)}

will try to fill in...

@@

Bioinformatics (from Tom Slezak) (source not in this doc): want to compare O(1000s) of terabye-scale data sets (metagenomic) with each other.  Datasets are distributed around the world.  Solution to the problem is not clear.

Climate (from Dan/Don) (\S\ref{climateDan}): Data (output of climate simulations) is distributed in multiple stores.  Current data analysis and visualization tools require the user to bring the data they want to analyze to a local system, then run the tools locally.  New tools will include the capability to add data transfer.  This will allow the set of tools to iterate - transfer data, do analysis, etc.

Astro (from Adam) (\S\ref{astroAdam}): Data taken by telescope - quick analysis done at telescope site for interesting (urgent) events (which may involve comparing new data with previous data).
Can get more data from other observatories if needed, request other observatories to take more data, or call a human. Data then transferred to archive site, which may be the observatory. At archive site, data analyzed , reduced, classified (some of this may be farmed out to grid resources).  Detailed analysis of new data vs. archived data done.  Reanalysis of all data done periodically.  Data stored in files and databases.

@@

@@

Astro (from Julian) (\S\ref{astroJulian}): Application builds a sky map.  Reads data from detectors, and gets data from simulations performed on-the-fly.

@@

Fusion (from Scott Klasky) (source not in this doc): multiple physics simulation codes run concurrently on a distributed set of parallel computers.  Data from some codes are streamed to other codes to link them into a single simulation.  Data is transformed in flight into needed inputs.  Data is also examined in situ to understand the overall state of the simulation.  Some data is stored for later data mining and visualization.

SOA Astro (from Adam) (\S\ref{astroSOAAdam}): services are orchestrated through a pipeline, including a data retrieval service.  Data moved through the pipeline, intermediate and final products can be stored in AstroGrid storage service.

\subsection{Additional sources}

\katznote{discuss this at April workshop}

>>>>

\footnotesize{\author{Editors: J.D. Blower, Neil Chue-Hong, Simon Dobson, \\[-0.1em]
  Shantenu Jha, Daniel S. Katz, Omer Rana}}

@@

\begin{abstract}
\footnotesize{Many problems at the forefront of science, engineering, medicine,
  and the social sciences, are increasingly complex and
  interdisciplinary due to the plethora of data sources and
  computational methods available today.  A common feature across many
  of these problem domains is the amount and diversity of data and
  computation that must be integrated to yield insights. Such data is
  increasingly large-scale and distributed;
  the data-sets, their location and scheduling decision
  are time-dependent. % Applications need access
  % to these disparate data-sets and to the methods that operate on
  % them.  Data-intensive applications need first-class support for
  % distributed heterogeneous platform execution.
  The 3DPAS Research theme (funded by the eSI, Edinburgh and US NSF),
  aims to investigate applications that have these characteristics, in
  other words it operates at the triple point of {\it dynamic} and
  {\it distributed} and {\it data-intensive}, (3D) attributes.  This
  evolving document represents work in progress of the 3DPAS Research
  Theme.}
\end{abstract}

@@

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[scale=0.33]{figures/3dpas-wordle.pdf}
    \caption{Wordle of 3DPAS document}
    \label{Fig:wordle}
  \end{center}
\end{figure*}

@@

@@

Why bother about distributed, dynamic and data-intensive applications?

@@

\subsubsection{Types of Dynamic Data}

\begin{enumerate}
\item Streaming data -- data comes and goes
  \url{http://traffic.berkeley.edu/} and \url{http://lagrange.ce.berkeley.edu/fsn/}
\item Data set that is being operating upon is changing
\item Data has to be transformed
\item Placement decisions time-dependent
\end{enumerate}

@@

\subsubsection{Dynamic Data in a Web Environment}\label{sec:dynamicdataweb}
\emph{Just some placeholder notes from Jon.  Could belong elsewhere.}

@@

@@

\subsubsection{Answers from Steve}

@@

\subsection{Astrophysics -- Virtual Observatory
(Bob) \label{astroBob}}
\begin{itemize}
\item Analysis of data to generate event streams $\rightarrow$ lead to
the positioning of telescopes
\end{itemize}

@@

I assume that the ``application'' in question is the sum of the two
activities, despite the fact that they are different in many respects,
and I try and distinguish between them where appropriate.

@@

For the sea mammals the animal crawls up a beach, the sensor packages
sees the cellular network and transmits. For environmental missions,
the answer is unclear.

@@

Generally it's just a matter of
capturing at whatever resolution device the sensor pack can support.

@@

{\bf CDAT}: The Climate Data Analysis Tools (CDAT)~\cite{CDAT} are a
scriptable (Python) set of packages and tools that are currently
heavily used for analysis of climate data.  CDAT has been developed by
LLNL's Program for Climate Model Diagnosis and Intercomparison
(PCMDI), and is often used with data from the Earth System Grid
(ESG)~\cite{ESG}.  The most common current usage of these tools with
the data is that a user downloads a set of data from ESG to a local
resource, then uses the CDAT to analyze the data.  However, the PCMDI
group is now planning to support a new usage mode, where data transfer
can be scripted.  Initial development of this capability is planned
for February 2011, using the Globus online data transfer
service~\cite{globusOnline} as a scriptable component.

\subsection{Oceanographic/BADC Applications \label{oceanStephen}}
Climate from BADC + Oceanographic data (Data Nodes + Gateways)
(Stephen Pascoe)
\begin{itemize}
\item Dynamic aspects not clear -- primarily providing a data storage
(curated) system
\item Multiple sources of data -- and potentially coupling between
simulation and data
\item Issue of placement and scheduling of the data -- through the use
of federated registries
\item Identify how this links it to the ``application genie''
\end{itemize}

\subsection{Interactive exploration of environmental data
(Jon) \label{envJon}}

(Here's a synopsis of the chapter I'm writing for Malcolm's book on Data-Intensive Research.
I'm not sure how to merge this with the answers to the pro-forma questions below.  Also,
many of these sentences are cut-and-pasted from the book chapter - future copyright issue?)

In the environmental sciences the use of visualization techniques is vital for understanding
the ever-increasing volume and diversity of data that is being produced by Earth observing
systems and computer simulations.  The primary purpose of visualization is to gain insight
\cite{card_readings_1999}, but the majority of current tools in wide use are not adequate
for achieving this.  The time spent to create visualizations is too great, usually due to the
crippling overhead of low-level technical tasks such as file-format and metadata interpretation.
Most scientific tools generate static plots, which are shared in documents or emails, but 
neither the originating scientist nor the recipient can easily customize or adapt the visualization
to reveal new information.  The context of how the visualization was generated is usually lost
in the publication/sharing process, making interpretation and reproduction much more difficult.

Visualization of a single dataset is therefore more tedious and less useful than it should be.
These problems are compounded if the scientist wishes to examine multiple datasets simultaneously.
This is an extremely common task in environmental science, which forms an important part of model validation,
``ground truthing'', quality control and data assimilation.

There is much current interest in developing new, interactive modes of environmental data exploration and visualization, which can
play a key role at all stages of scientific investigation.   We consider that, for the case of interactive
data exploration, concerns of performance,
intuitiveness and user engagement tend to outweigh the desire for sophistication in data processing. 
We therefore argue that a new generation of "4D GIS" tools, based on the principles of simplicity,
open standards and user-friendliness is required to gain greater insight from the environmental data
deluge.  These new tools will not replace existing approaches, but rather complement them.

Increasingly, environmental data are distributed in multiple locations, and are frequently updated
(often several times a day).  Interactive exploration of such large, dynamic, distributed data is therefore
a strong technical challenge.  Data infrastructures must support the ability to quickly extract
subsets of large data, and any data processing must happen quickly, with the minimum of overhead.
Web services are therefore playing an increasingly-important role, and there is much interest in
cloud technologies to provide the necessary scalability at the back end.  Grid technologies, being essentially batch-mode
systems, are less suitable for this kind of application as the scheduling overhead tends to dominate 
the processing time for simple jobs, seriously hampering interactivity.

Performance is therefore a key challenge, exacerbated by the dynamic nature of the data, which makes
it hard to achieve scalability through caching (see section~\ref{sec:dynamicdataweb}).  [Note: could probably
merge material from that section here.]  Security is another serious concern: many environmental datasets
are held under access control.  In a distributed environment, it is common for machines to access data
on behalf of end users (e.g.\ a processing service might download its input data from a remote store).
Finding a means for the user to delegate his/her authority to a multi-web-service infrastructure is
a key current challenge.  The MashMyData project (http://www.mashmydata.org) is investigating two solutions, based
respectively on Grid proxy certificates and OAuth.

@@

The applications often take the form of graphical tools that share a
lot in common with Geographic Information Systems (GIS). However these
applications are much better suited to scientific data, which are
large and multidimensional.  The general “vision” is of a map-based
interface, onto which different datasets can be overlain.  A
processing step may be initiated by rubber-banding an area of the map
and selecting from a list of algorithms that process data within the
selected area, perhaps calculating statistics of the data.

@@

Usually the applications are driven by data served through web
services of various kinds.  These web services may be “fed” by data
coming from instruments or from computer simulations.  A key challenge
is how to integrate compute services with these data services in an
efficient, easy-to-use and transparent way, allowing fast data
processing.  The processing algorithms are generally quite simple and
so it is frustrating to succumb to the high overhead of some
scheduling systems.  The infrastructure probably has more in common
with the Web (or perhaps cloud) than the Grid.

@@

Dynamic data are only sometimes used, but may arise from “real-time”
feeds from instruments or from looking at “live” results from a model.
Data from instruments are “small” 1MB-100MB but model output may be
very big (GB-TB).

@@

@@

Bioinformatics: want to compare O(1000s) of terabye-scale data sets (metagenomic) with each other.  Datasets are distributed around the world.  Solution to the problem is not clear.

Climate (from Dan) (\S\ref{climateDan}): Data (output of climate simulations) is distributed in multiple stores.  Current data analysis and visualization tools require the user to bring the data they want to analyze to a local system, then run the tools locally.  New tools will include the capability to add data transfer.  This will allow the set of tools to iterate - transfer data, do analysis, etc.

Astro (LSST): Data taken by telescope - quick analysis done at telescope site for interesting (urgent) events (which may involve comparing new data with previous data). Data then transferred to archive site. At archive site, data analyzed, reduced, classified.  Detailed analysis of new data vs. archived data done.  Reanalysis of all data done periodically.  Data stored in files and databases.

@@

Stephen Pascoe's Oceanographic app (\S\ref{oceanStephen}):
\katznote{need summary here - but not sure what to write...}

@@

Astro (CMB): Application builds a sky map.  Reads data from detectors, and gets data from simulations performed on-the-fly.

@@

Fusion: multiple physics simulation codes run concurrently on a distributed set of parallel computers.  Data from some codes are streamed to other codes to link them into a single simulation.  Data is transformed in flight into needed inputs.  Data is also examined in situ to understand the overall state of the simulation.  Some data is stored for later data mining and visualization.
